% ============================================================
% RLHF on a Shoestring — project report (paper-style, two-column)
% Compile: pdfLaTeX (Overleaf: Menu -> Compiler -> pdfLaTeX)
% Layout: full-width title + abstract, two-column body (like NeurIPS/ACL preprints).
% Content mirrors report.md (Vietnamese working notes) — keep the two in sync.
% If your university requires its own template: move the \section bodies into it,
% keep references.bib as-is.
% ============================================================
\documentclass[10pt,twocolumn]{article}

\usepackage[utf8]{inputenc}
\usepackage[T5]{fontenc}         % T5 renders both English and Vietnamese diacritics
\usepackage[english]{babel}
\usepackage[a4paper,margin=2cm,columnsep=0.7cm]{geometry}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{xcolor}

% Placeholder for pending experimental numbers — search for \tbd to see what's missing
\newcommand{\tbd}[1]{\textcolor{red}{[TBD: #1]}}

\title{\textbf{RLHF on a Shoestring: Building a Vietnamese Chatbot\\ from a Raw Base Model on Free-Tier GPUs}}
\author{Dinh Phuc\thanks{\tbd{full name, student ID, department/university, advisor}}}
\date{\today}

\begin{document}

% Full-width title + abstract, then two-column body — standard paper layout
\makeatletter
\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
\noindent Most publicly deployed chatbots are produced by a multi-stage alignment pipeline --- continual pretraining, supervised fine-tuning, reward modeling, and preference optimization --- yet nearly all accessible accounts of this pipeline come from industrial labs operating hundreds of GPUs, leaving open the question of whether the full recipe can be executed, studied, and measured end-to-end under a student-scale compute budget. This report answers that question affirmatively for Vietnamese. Starting from a raw, non-instruction-tuned base model (\texttt{Qwen/Qwen3-1.7B-Base}), we build and operate the complete RLHF pipeline: continual pretraining (CPT) on roughly 400M tokens of quality-filtered Vietnamese text (Wikipedia and FineWeb2) with a 20\% English replay mixture to counter catastrophic forgetting; supervised fine-tuning on 12.7k multi-turn Vietnamese conversations with response-only loss masking; reward model training on human preference pairs; direct preference optimization (DPO) as the main alignment path with PPO as a multi-GPU ablation; quantitative evaluation; and deployment via GGUF quantization to an Ollama/Open WebUI chat interface. Every training stage runs on free-tier hardware --- a single Kaggle T4 (16\,GB, 30\,h/week quota, sessions capped at $\sim$9\,h) --- made possible by a checkpoint-resume mechanism that pushes full trainer state (adapter, optimizer, scheduler, RNG) to the HuggingFace Hub at every save and stops itself before the session budget expires, so that any stage can be split across an arbitrary number of sessions with exact continuation. Beyond the artifact itself, the project is designed around three single-variable ablations that budget-constrained practitioners actually face: (E1) whether CPT measurably improves a downstream Vietnamese assistant relative to fine-tuning the raw base directly, measured on VMLU and LLM-judged win-rate; (E2) how closely single-GPU DPO matches multi-GPU PPO in quality per GPU-hour on identical preference data; and (E3) whether a 20\% replay mixture prevents forgetting, tracked by parallel Vietnamese/English held-out losses over the full CPT run. \tbd{one or two sentences summarizing the main quantitative findings once E1--E3 complete}. All notebooks, datasets, and intermediate checkpoints are public or reproducible from public sources, and every result in this report can be regenerated from a free Kaggle account.
    \end{abstract}
    \vspace{1.5em}
  \end{@twocolumnfalse}
]
\makeatother

\section{Introduction}

Most student-level ``Vietnamese chatbot'' projects follow the same route: take an instruction-tuned model off the shelf (Llama-Instruct, Qwen-Instruct), fine-tune it lightly, and demo the result. That route ships a product quickly, but it skips the entire part worth studying: \emph{where does the ability to chat come from?} This project takes the opposite route --- starting from raw weights that can only predict the next token, we build every layer that turns them into a Vietnamese assistant, following the industrial recipe established by InstructGPT~\citep{ouyang2022instructgpt} and applied to South-East Asian languages by Sailor2~\citep{dou2025sailor2}.

We impose one additional constraint on ourselves: \textbf{free compute only}. All training runs on a single Kaggle T4 (16\,GB VRAM, 30\,h/week quota, sessions capped at roughly nine hours). This constraint is not a weakness to hide but part of the problem statement: every technique in this project must survive an environment where the training session can be terminated at any time.

\section{Contributions}

The contribution of this project is not a model that beats existing Vietnamese LLMs --- impossible at student-scale resources --- but the following three items:

\begin{enumerate}[leftmargin=*]
  \item \textbf{A complete pipeline from scratch.} We build and operate the full chain CPT $\to$ SFT $\to$ RM $\to$ DPO $\to$ Eval $\to$ Deploy from a raw base model, rather than fine-tuning an already instruction-tuned checkpoint as most comparable projects do. Every stage ships as an independent, actually-runnable notebook with a smoke test.
  \item \textbf{Three design questions answered with numbers} (Section~\ref{sec:experiments}) --- questions that large-lab publications rarely address at small scale: (i) is continual pretraining worth its cost for Vietnamese; (ii) how far does DPO substitute for PPO, and at what cost; (iii) does a 20\% English replay mixture actually stop catastrophic forgetting.
  \item \textbf{A fully reproducible recipe on free infrastructure.} A checkpoint-resume mechanism over the HuggingFace Hub (pushing the entire checkpoint folder --- adapter, optimizer, scheduler, RNG state --- at every \texttt{save\_steps}; stopping itself before the session budget; the next session resumes at the exact step by re-running the notebook) turns every training stage into an indefinitely repeatable chain of 8-hour sessions. Sailor2 published a cookbook for teams with hundreds of GPUs~\citep{dou2025sailor2}; this project is the cookbook for someone with one borrowed T4.
\end{enumerate}

Sailor2 serves as a \emph{reference point} in evaluation, not as a starting checkpoint.

\section{Related Work}

\paragraph{The relative roles of pretraining and alignment.}
LIMA~\citep{zhou2023lima} showed that most knowledge and linguistic competence is acquired during pretraining, while alignment mainly teaches \emph{format and style} (the superficial alignment hypothesis); URIAL~\citep{lin2023urial} confirmed this quantitatively at the token level. This is the theoretical basis for keeping the CPT stage rather than relying on conversational data downstream. On CPT technique, \citet{ibrahim2024simple} showed that learning-rate re-warming/re-decaying combined with replay of old data matches full retraining; \citet{gururangan2020dont} is the classic result on domain-adaptive pretraining; \citet{luo2023forgetting} measured catastrophic forgetting and found smaller models forget more --- directly relevant at our 1.7B scale.

\paragraph{The standard recipe for regional languages.}
Every strong Vietnamese or South-East Asian model applies CPT on native-language corpora before alignment, without exception: Sailor~\citep{dou2024sailor} and Sailor2~\citep{dou2025sailor2} (400B SEA tokens + 100B replay from Qwen2.5), SeaLLMs~\citep{nguyen2023seallms}, VinaLLaMA~\citep{nguyen2023vinallama} (800B tokens of CPT on Llama-2), and PhoGPT~\citep{nguyen2023phogpt} (pretrained from scratch on 102B tokens). Our base model is Qwen3~\citep{qwen3report}.

\paragraph{RLHF and its lighter substitutes.}
The SFT $\to$ RM $\to$ PPO chain was established by \citet{christiano2017preferences,stiennon2020summarize,ouyang2022instructgpt}, with the deepest empirical treatment of reward models in \citet{bai2022hh}; PPO~\citep{schulman2017ppo} is the underlying RL algorithm, and OpenRLHF~\citep{hu2024openrlhf} is the framework we use for the RM stage and the PPO ablation. DPO~\citep{rafailov2023dpo} collapses RLHF into a classification loss requiring only a policy and a reference model --- the reason it is our main path on a single GPU.

\paragraph{What makes a T4 sufficient.}
LoRA~\citep{hu2021lora}, QLoRA~\citep{dettmers2023qlora} (NF4 4-bit base + fp16 adapters + paged optimizers), and rsLoRA~\citep{kalajdzievski2023rslora} ($\alpha/\sqrt{r}$ scaling for high ranks) are the trio that fits 1.7B-parameter training into 16\,GB of VRAM. The Vietnamese corpus comes from FineWeb2~\citep{penedo2025fineweb2}; the English replay from FineWeb-Edu~\citep{penedo2024fineweb}.

\section{Pipeline}
\label{sec:pipeline}

Detailed per-stage design lives in the repository spec (Section 6); runnable notebooks in \texttt{notebooks/}. Table~\ref{tab:pipeline} summarizes.

\begin{table*}[t]
\centering\small
\caption{The six-stage pipeline, all starting from raw \texttt{Qwen3-1.7B-Base}.}
\label{tab:pipeline}
\begin{tabularx}{\textwidth}{@{}lXl@{}}
\toprule
\textbf{Stage} & \textbf{Content} & \textbf{Output (HF Hub)} \\
\midrule
1. CPT & QLoRA $r{=}64$, $\sim$400M tokens (Wikipedia-vi + FineWeb2 \texttt{vie\_Latn} + 20\% FineWeb-Edu replay), 6000 steps & \texttt{Qwen3-1.7B-vi-cpt} \\
2. SFT & QLoRA $r{=}16$, 12.7k multi-turn conversations, ChatML template, loss on assistant turns only & \texttt{Qwen3-1.7B-vi-sft} \\
3. RM & OpenRLHF \texttt{train\_rm} on Vi-Alpaca preference pairs & \texttt{Qwen3-1.7B-vi-rm} \\
4. DPO & TRL \texttt{DPOTrainer} (main path); PPO on Modal (ablation) & \texttt{Qwen3-1.7B-vi-dpo} \\
5. Eval & VMLU + LLM-judge win-rate + RM score on held-out split & --- \\
6. Deploy & GGUF $\to$ Ollama + Open WebUI & \texttt{...-dpo-gguf} \\
\bottomrule
\end{tabularx}
\end{table*}

\paragraph{Why Qwen3 rather than Llama.}
Two reasons. First, the tokenizer: Qwen's $\sim$152k-entry vocabulary, trained on multilingual data, gives Vietnamese a fertility (tokens per word) close to English, whereas Llama-2's 32k vocabulary shatters diacritic-bearing syllables into byte fragments --- and the tokenizer is the one thing CPT cannot fix without vocabulary extension as in SeaLLMs~\citep{nguyen2023seallms}. Second, budget arithmetic: with a fixed token budget, a high starting point plus a light top-up travels further than a low starting point plus a massive dose --- VinaLLaMA needed 800B tokens to bring Llama-2 up to standard in Vietnamese~\citep{nguyen2023vinallama}, roughly 2000$\times$ our entire budget.

\section{Experimental Design}
\label{sec:experiments}

Three experiments, each varying exactly one factor (Table~\ref{tab:ablations}). Evaluation rules: the held-out split is never used to train RM/DPO/PPO; the LLM judge scores each pair twice with answer positions swapped to cancel position bias~\citep{zheng2023judge}; VMLU~\citep{vmlu2025} measures Vietnamese knowledge; Sailor2-1B serves as an external reference point.

\begin{table*}[t]
\centering\small
\caption{The three ablation questions and their metrics.}
\label{tab:ablations}
\begin{tabularx}{\textwidth}{@{}lXXl@{}}
\toprule
\# & \textbf{Question} & \textbf{Comparison} & \textbf{Metrics} \\
\midrule
E1 & Is CPT worth it? & SFT-from-CPT vs.\ SFT-from-raw-base, all else equal & VMLU, win-rate \\
E2 & Does DPO substitute for PPO? & DPO (single T4) vs.\ PPO (Modal, multi-GPU), same preference data & win-rate, RM score, GPU-hours \\
E3 & Does replay stop forgetting? & \texttt{eval\_vi\_loss} vs.\ \texttt{eval\_en\_loss} across all 6000 CPT steps & loss curves \\
\bottomrule
\end{tabularx}
\end{table*}

\section{Results}
\label{sec:results}

\tbd{Fill in as stages complete --- one table of numbers plus one paragraph of commentary per stage. Data sources: \texttt{trainer\_state.json} inside the Hub checkpoints (CPT/SFT logs), and the stage-5 evaluation scripts.}

\subsection{E1 --- The value of CPT} \tbd{VMLU + win-rate table}
\subsection{E2 --- DPO vs.\ PPO} \tbd{quality + GPU-hour table}
\subsection{E3 --- Catastrophic forgetting} \tbd{vi/en loss-curve figure from CPT logs}

\section{Anticipated Objections}

\begin{description}[leftmargin=0pt,style=nextline]
  \item[``What does yours add over existing models?''] We do not compete on quality with models trained on hundreds of GPUs --- we reproduce \emph{how those models are made}, at student scale, measuring every layer. The added value is Table~\ref{tab:ablations}: numbers that using an off-the-shelf model can never produce.
  \item[``Why not just use Sailor2?''] With a ready-made model there is no layer left to measure --- no research question can be answered. Sailor2 does appear in this project, in its proper role: as a reference point in evaluation.
  \item[``Your model loses to ChatGPT --- so what is it for?''] Correct, and winning was never the goal --- controlling every layer of the process is. ChatGPT does not show you its reward model's loss curve.
  \item[``Why such a small model (1.7B)?''] Because our questions are questions of \emph{method}, not \emph{scale} --- every ablation conclusion is measurable at 1.7B. Scaling up is a question of money, not of science.
  \item[``Is your CPT actually useful, or just following the trend?''] We make no bare claim --- E1 exists precisely to answer this with numbers. If the delta turns out small, that too is an honest, reported finding.
  \item[``Can work on free Kaggle be taken seriously?''] The compute constraint is part of the problem statement: every training stage checkpoints and resumes through the Hub and is 100\% reproducible from public notebooks. Reproducibility is something many large publications still lack.
\end{description}

\section{Reproducibility Statement}

All notebooks run directly on Kaggle's free T4 tier; datasets and intermediate checkpoints live on the HuggingFace Hub under the naming convention \texttt{<user>/Qwen3-1.7B-vi-\{cpt,sft,rm,dpo\}}; all seeds are fixed (42); complete training logs are preserved in each checkpoint's \texttt{trainer\_state.json}. Anyone with a Kaggle account and an HF token can reproduce every stage by re-running the notebooks.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
